25 Years of Water Quality Change in Rhode Island Lakes and Ponds

### 2018-11-05 manuscript/manuscript.Rmd
drafted some z-score and tred text

There are many different methods for analyzing time series data for trends.  One of the difficulties that is encountered when you have multiple sampling locations is how to identify a trend when you have variation at the scale of the sampling location as well as variation in when a sampling location might have been addedd to a dataset.  For instance, if you have long-term data on water quality for several ponds, yet the cleaner ponds sampled more frequently in early years then a simple comparison of raw-values over time might show a decrease in water quality.  Thus, it is necessary to account for this type of site-specific variation.  This is very similar to how the long-term temperature trends have been calculated by the XXX [FILL IN GROUP DOING THIS]

@@

>>>>

There are many different methods for analyzing time series data for trends.  One of the difficulties that is encountered when you have multiple sampling locations is how to identify a trend when you have variation at the scale of the sampling location as well as variation in when a sampling location might have been addedd to a dataset.  For instance, if you have long-term data on water quality for several ponds, yet the cleaner ponds sampled more frequently in early years then a simple comparison of raw-values over time might show a decrease in water quality.  Thus, it is necessary to account for this type of site-specific variation.  This is very similar to how the long-term temperature trends have been calculated by the XXX [FILL IN GROUP DOING THIS].  The general approach is to calculate site-specific deviations from a pre-determined reference period.  This allows all sites to be shifted to a common base line and the deviations, or anomalies, show change relative to the referenc period.

@@

The anomaly approach is very useful and is the approach we chose to follow.  However, using anomalies works best with a single measure (e.g temperature) or with mulitple measurements that are on the same scale.  The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.  For instance, temperature anomalies will like range +/- 2 to 3 degrees, whereas phosphorus anomalies, might range +/- XX to XX ug/l.  To standardize these values we used the common approach of dividing each anomlaly by the standard deviation for the reference period.  This value is commonly referred to as a Z-score.  We use these z-scores to examine each water quality variable for a trend over the time period of 1990 to 2016.
